Tools for identifying implausible counts

### 2019-04-07 Rmd/count_check_MS.Rmd
line returns after every sentence

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms. This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001]. However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum. There is a documented case of mis-reported count sums: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen. Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.

@@

<!-- This paper develops a method, related to GRIM [@Brown2017], that can test if the count sums are probably as large as reported. -->

This paper aims to test the methods presented above, and discusses some complications and caveats.

<!-- estimates the prevalence of microfossil assemblages without singletons, and develops additional tests for samples without singletons. Next, these tests are applied to percent data where the count sums are known, and then to percent data where the count sums are not known. -->

@@

Data sets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources. A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness. Datasets with possible data mishandling are anonymised. This paper does not attempt to be an exhaustive survey of all the data available.

@@

The third dataset consists of `r nrow(last_chance_meta)` chironomid head capsule counts from Last Chance Lake [@Axford2017]. It includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked. The percent are given to two decimal places: to account for rounding errors, I use the smallest and largest values consistent with the reported percentage and show this range as a error bar. All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)). The factor of two difference is because in some counts the rarest taxon is represented by a half head capsule. Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers), occasionally other fractions are reported. Fractional counts make the estimated count too high, so do no risk incorretly flagging counts as being too small.

@@

Dataset 3 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy. The associated paper reports that count sums are 200 (excluding fern spores). Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english()` of the counts has a singleton. The other counts have minimum abundances of two or four pollen grains. With the exception of the single singleton, all `r pollen1_summary$individual$n %>% invisible()` species counts are divisible by two`r pollen1_summary$individual$not_two %>% insist(. == 1) %>% invisible()`. The probablility that these counts reflects the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.

>>>>

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often 50 for chironomids and several hundred for pollen and diatoms. 
This is important because larger count sums are associated with smaller uncertainties, both is in relative abundance of taxa and derived statistics such are transfer function reconstructions [@Heiri2001]. 
However, given the time-consuming nature of microfossil counting, especially when preservation is poor or concentrations are low, there is an incentive to misreport the minimum count sum. 
The risk that count sums will be mis-reported is not just theoretical: Larocque-Tobler et al [@Larocque-Tobler2015] reported that their chironomid count sums were at least 50 head capsules, a subsequent corrigendum  [@Larocque-Tobler2016] acknowledged that count sums were actually as low as nineteen. 
Telford [-@Telford2019] reports some other cases where count sums may be much lower than reported.

@@

This paper aims to test the methods presented above and present some cases with unexpected results. 
Some complications and caveats are discussed.

@@

Datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea and other sources. 
A range of ecological and palaeoecological data were sought to allow for difference in typical count sum and species richness. 
Datasets with possible data mis-reporting are anonymised, but no attempt is made to diagnose whether error or misconduct are responsible.
This paper does not attempt to be an exhaustive survey of all the data available.

@@

The third dataset consists of `r nrow(last_chance_meta)` chironomid head capsule counts from Last Chance Lake [@Axford2017]. 
It includes both the count sums and the percentage of each taxa, allowing the count sums estimated from the rarest taxon to be checked. 
The percent are given to two decimal places: to account for rounding errors, I use the smallest and largest values consistent with the reported percentage and show this range as a error bar. 
All the estimated count sums are, within error, either identical to the reported count sum or twice as much (Fig. \@ref(fig:last-chance-plot)). 
The factor of two difference is because in some counts the rarest taxon is represented by a half head capsule. 
Reporting half microfossils is common for several microfossil groups, including chironomids and pollen (especially for bisaccate conifers), occasionally other fractions are reported.
Fractional counts make the estimated count too high, so do no risk incorrectly flagging counts as being too small.

@@

Dataset 4 is a `r n_groups(readd(pollen1_counts))` count pollen stratigraphy.
The associated paper reports that count sums are 200 (excluding fern and fungal spores).
The archived data have counts sums between `r pollen1_summary$count$min_count` and `r pollen1_summary$count$max2_count`, with the exception of one count of `r pollen1_summary$count$max_count` (which may be so high because of a possible data entry error).
However, these counts sums include trilete spores, probably from a fern or fern ally, with an abundance as high as `r pollen1_summary$max_trilete`.
Unexpectedly, only `r pollen1_summary$count$has_singletons %>% insist(. == 1) %>% as.english() %>% as.character()` of the counts has a singleton.
The other counts have minimum abundances of two or four pollen grains.
With the exception of the single singleton, all `r pollen1_summary$individual$n %>% invisible()` species counts are divisible by two`r pollen1_summary$individual$not_two %>% insist(. == 1) %>% invisible()`.
The probability that these counts reflects the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled in some way.
